Hydrodynamics of a Semipermeable Vesicle Under Flow and Confinement

### 2021-01-21 vesicle_code/docs/permeableVes/main.tex
yny

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Formulation}
\begin{figure*}[hbp]
  \centering
  \ifTikz
  \input{figures/relaxationComposite.tikz}
  \else
  \includegraphics{figures/relaxationComposite.pdf}
  \fi
  \caption{\label{fig:relaxationComposite} (a) Snapshots of four
  semipermeable vesicles with $\beta = 10^{-7}$ submerged in a fluid
  with no background flow flow. Time goes from left to right. The red
  regions correspond to influx and the blue regions correspond to
  efflux. In all cases, the vesicle reaches a steady state circular
  shape. (b) The reduced area of each of the simulations. The particular
  snapshots in (a) occur at the marks along the curve.}
\end{figure*}
We consider a two-dimensional semipermeable vesicle (permeable only to
water) suspended in a viscous fluid (Figure~\ref{fig:sketch}), and we
let $\gamma$ denote the membrane boundary. We let
$\alpha=4\pi A/L^2$ be the vesicle's reduced area where $A$ is its area
and $L$ is its length. To focus on the effect of water flow on vesicle
hydrodynamics in various conditions, we assume the fluid is the same
inside and outside the vesicle membrane. On the membrane the permeable
water flux gives rise to a difference between the fluid velocity
${\uu}({\xx}\in \gamma)$ and the membrane velocity
$\dot{\xx}$~\cite{yao-mor2017}:
\begin{align}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\nn$ is the outward normal (Figure~\ref{fig:sketch}), $k_w$ is
the hydraulic permeability, $R$ is the ideal gas constant, $T$ is the
temperature (K), $\triangle c$ is contrast in a solute concentration
(mol/volume), and $\ff_\mathrm{mem}$ is the membrane stress that
consists of both bending $ \ff_\mathrm{ben}= -k_b \xx_{ssss}$ and
tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$ ($s$ is the arclength and
$\sigma$ is the membrane tension along the membrane, see Supplementary
Material). Using $k_w\sim 10^{-12}$ m$^2$s/kg (typical of mammalian cell
membranes~\cite{LiMoriSun2015_PRL}) and an osmolarity $\triangle c =
0.1$ mol/L (consistent with the osmotic forces in vesicle aspiration
experiments~\cite{OlbrichRawiczNeedhamEtAl2000_BJ}), the corresponding
water flow is of the order $v_\mathrm{o}=0.1 \; \mu$m/s. Assuming a
membrane bending stiffness of $k_b = 10^{-19}$ J, a membrane tension of
$10^{-3}$ N/m (one tenth of the lysis tension), and a characteristic
length scale of $10^{-6}$ m, the tension contribution ($v_\mathrm{t}$)
and elastic stress contribution ($v_\mathrm{e}$) to the permeable water
flow can be scaled to $v_\mathrm{o}$ as
$v_\mathrm{o}:v_\mathrm{t}:v_\mathrm{e}=10^{0}:10^{-2}:10^{-6}$.
Furthermore, we note that in the absence of an osmotic stress, the water
permeable flow can be significant when the membrane tension is large due
to the external stress such as the hydrodynamic stress due to an
external flow or confinement. In the following we explore both.

@@

In experiments, vesicles under shear flows are measured for a minute
(chapter 19 in~\cite{Dimova2020_GVB}), a duration over which a vesicle
can be assumed impermeable~\cite{AbkarianViallat2005_BJ}. Our
simulations of a semipermeable vesicle in these flows show that water
permeability may lead to different vesicle hydrodynamics over a time
scale of 10 to 100 minutes. For vesicles in a narrow microfluidic
channel, the experiments usually do not last longer than a few
minutes~\cite{abk-fai-sto2006}. Our simulations show that strong
confinement can lead to amplification of water permeability at such
short time scales.

@@

We consider a semipermeable vesicle in a planar shear flow,
characterized by the elastic capillary number $Ca_\mathrm{E}=\chi\tau$,
where $\tau$ is the characteristic time scale in the
non-dimensionalization (see Supplementary Material) and $\chi$ is the
shear rate of the planar shear flow $\uu_{\infty}(\xx) = \chi (y,0)$.
When $\beta=0$ (impermeable case), a vesicle's reduced area and
viscosity contrast determine whether it tank treads or
tumbles~\cite{fin-lam-sei-gom2008, kra-win-sei-lip1996}. Focusing on
semipermeable vesicles with no viscosity contrast, we find that a
semipermeable vesicle tilts to an inclination angle and undergoes
tank-treading dynamics, similar to the case of an impermeable vesicle.

In Figure~\ref{fig:shearComposite}(a), using four initial reduced areas
and two flow rates, we plot the reduced area of a semipermeable vesicle
with $\beta = 10^{-3}$. As observed in a quiescent flow, the final
vesicle shape is independent of the initial reduced area. However, it
does depend on the flow rate. We also observe that the reduced area
remains constant until $t\sim 10$ minutes, when the vesicle starts to
tank-tread and the normal component of the force-balance velocity
becomes significantly smaller than the permeable velocity, leading to a
changing reduced area.

For a variety of semipermeability rates and $Ca_\mathrm{E}$, a
semipermeable vesicle reaches a steady equilibrium as in
Figure~\ref{fig:shearComposite}(a). Since the final shape is independent
of the initial reduced area, we initialize the vesicle with reduced area
$\alpha = 0.65$ for all simulations in
Figure~\ref{fig:shearComposite}(b), where we summarize the steady state
vesicle shape and its reduced area. We see that the capillary number
(flow rate) significantly affects the final vesicle shape, but the
semipermeability rate has a minor effect. Instead, as we saw in the
quiescent analysis, $\beta$ sets a time scale since the time required
for a vesicle to reach its steady state configuration is proportional to
$\beta^{-1}$.

@@

At the beginning of the simulations a semipermeable vesicle is
initialized above the center line $y=0$, and it migrates towards the
center line as in the impermeable case~\cite{dan-vla-mis2009}. 
%
%However,
%for an impermeable vesicle, depending on the initial reduced area and
%the maximum flow velocity $U$, the equilibrium vesicle shape is either
%an axisymmetric parachute or bullet, or an asymmetric tank-treading
%slipper~\cite{kao-bir-mis2009}. 
%
%To validate our methods,
%in Figure~\ref{fig:parabolicComposite}(a), we construct a phase diagram
%for the steady state shape of an impermeable vesicle
%vesicle~\cite{kao-bir-mis2009} (see Figure 2). While the vesicle shapes
%qualitatively agree, Kaoui et al.~report that the vesicles with
%$\alpha=0.7$ and flow rates $U = 1200 \mu m/s$ and $U = 1600 \mu m/s$ are
%axisymmetric. In our simulations these vesicles are slightly asymmetric,
%are undergoing a tank treading motion, and we therefore report them as
%asymmetric slippers.
%
Using several initial reduced areas and flow rates, we recreate the
impermeable vesicle phase diagram for vesicles with $\beta = 10^{-3}$.
The equilibrium vesicle shape for different $U$ and initial reduced area
is summarized in Figure~\ref{fig:parabolicComposite}(a), and the
vertical displacement of the vesicle's center of mass is in
Figure~\ref{fig:parabolicComposite}(b). The effect of semipermeability
of a vesicle with $\alpha(0) = 0.9$ and $U=800\; \mu$m/s on its reduced
area and migration pattern are shown in
Figures~\ref{fig:parabolicComposite}(c)
and~\ref{fig:parabolicComposite}(d), respectively.

As observed for the planar shear flow, the steady state reduced area is
independent of the initial reduced area, and larger flow rates result in
smaller steady state reduced areas. We report the reduced area of the
equilibrium vesicle shape in Figure~\ref{fig:parabolicComposite}(a). At
the smallest flow rate of $U= 200\; \mu$m/s, the steady state reduced
area is large, and the result is an axisymmetric bullet. At the larger
flow rates, the steady state reduced area is smaller, and the result is
an asymmetric tank-treading slipper. We observe that the membrane
permeability to water drastically alters the vesicle shape and position
relative to the flow at equilibrium: At $\beta=10^{-3}$ the equilibrium
vesicle depends only on the maximum velocity $U$ of the nonlinear shear
flow. The higher $U$ the smaller the equilibrium reduced area and the
further away vesicle is relative to the flow center at $y=0$.

@@

Cells experience stretching and release messengers when they pass
through capillary vessels~\cite{Wan2008_PNAS, ForsythWan2011_PNAS,
Pak2015_PNAS}. Many cells exhibit extraordinary flexibility as they go
through narrow channels~\cite{AuStoreyMoore2016_PNAS}, and adjustment of
area to volume ratio may be a key factor for a successful passage.  Here
we quantify the effect of water permeability on a vesicle going through
a stenosis (a long microfluidic channel with a width smaller than the
cell radius).  Based on the findings in~\cite{abk-fai-sto2006}, we first
examine the effect of semipermeability on the pressure fluctuation
across a stenosed geometry. In this experiment, a pressure drop is used
to squeeze a red blood cell through a constriction with diameter
$5\;\mu$m and length $45\;\mu$m. We set up a similar computational
geometry with a Poiseuille flow imposed at the inlet and outlet
(Figure~\ref{fig:stenosisComposite}(a)). The flow velocity $U_{\max} =
1000 \;\mu$m/s is maintained at the inlet and outlet and is chosen so
that the residency time of the vesicle in the stenosis channel is $t_R
\approx 25$ ms which agrees with the experimental
results~\cite{abk-fai-sto2006}. When the red blood cell enters the
channel, an increase in the pressure drop is required to maintain a
constant flow rate, and the difference in the pressure drop is called
the excess pressure.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Formulation}
\begin{figure*}[hbp]
  \centering
  \ifTikz
  \input{figures/relaxationComposite.tikz}
  \else
  \includegraphics{figures/relaxationComposite.pdf}
  \fi
  \caption{\label{fig:relaxationComposite} (a) Snapshots of four
  semipermeable vesicles with $\beta = 10^{-7}$ submerged in a fluid
  with no background flow flow. Time goes from left to right. The red
  regions correspond to influx and the blue regions correspond to
  efflux. In all cases, the vesicle reaches a steady state circular
  shape. (b) The reduced area of each of the simulations. The particular
  snapshots in (a) occur at the marks along the curve.}
\end{figure*}
We consider a two-dimensional semipermeable vesicle (permeable only to
water) suspended in a viscous fluid (Figure~\ref{fig:sketch}), and we
let $\gamma$ denote the membrane boundary. We let
$\alpha=4\pi A/L^2$ be the vesicle's reduced area where $A$ is its area
and $L$ is its length. To focus on the effect of water flow on vesicle
hydrodynamics in various conditions, we assume the fluid is the same
inside and outside the vesicle membrane. On the membrane the permeable
water flux gives rise to a difference between the fluid velocity
${\uu}({\xx}\in \gamma)$ and the membrane velocity
$\dot{\xx}$~\cite{yao-mor2017}:
\begin{align}
  \uu - \dot{\xx} = - k_w (R T \triangle c + \ff_\mathrm{mem} \cdot \nn) \nn, \qquad
  \xx \in \gamma,
\end{align}
where $\nn$ is the outward normal (Figure~\ref{fig:sketch}), $k_w$ is
the hydraulic permeability, $R$ is the ideal gas constant, $T$ is the
temperature (K), $\triangle c$ is contrast in a solute concentration
(mol/volume), and $\ff_\mathrm{mem}$ is the membrane stress that
consists of both bending $ \ff_\mathrm{ben}= -k_b \xx_{ssss}$ and
tension $\ff_\mathrm{ten}=(\sigma \xx_s)_s$ ($s$ is the arclength and
$\sigma$ is the membrane tension along the membrane, see Supplementary
Material). Using $k_w\sim 10^{-12}$ m$^2$s/kg (typical of mammalian cell
membranes~\cite{LiMoriSun2015_PRL}) and an osmolarity $\triangle c =
0.1$ mol/L (consistent with the osmotic forces in vesicle aspiration
experiments~\cite{OlbrichRawiczNeedhamEtAl2000_BJ}), the corresponding
water flow is of the order $v_\mathrm{o}=0.1 \; \mu$m/s. Assuming a
membrane bending stiffness of $k_b = 10^{-19}$ J, a membrane tension of
$10^{-3}$ N/m (one tenth of the lysis tension), and a characteristic
length scale of $10^{-6}$ m, the tension contribution ($v_\mathrm{t}$)
and elastic stress contribution ($v_\mathrm{e}$) to the permeable water
flow can be scaled to $v_\mathrm{o}$ as
$v_\mathrm{o}:v_\mathrm{t}:v_\mathrm{e}=10^{0}:10^{-2}:10^{-6}$.
Furthermore, we note that in the absence of an osmotic stress, the water
permeable flow can be significant when the membrane tension is large due
to the external stress such as the hydrodynamic stress from an
external flow or confinement. In the following we explore both.

@@

In experiments, vesicles under shear flows are measured for a few minutes
(chapter 19 in~\cite{Dimova2020_GVB}), a duration over which a vesicle
can be assumed impermeable~\cite{AbkarianViallat2005_BJ}. Our
simulations of a semipermeable vesicle under these flows in open geometry show that water
permeability may lead to different vesicle hydrodynamics only after  a duration of $\sim 200$ minutes. For vesicles in a narrow microfluidic
channel, the experiments usually do not last longer than a few
minutes~\cite{abk-fai-sto2006}. Our simulations show that strong
confinement can lead to amplification of water permeation at a time scale shorter than a few minutes.
%short time scales.

@@

We consider a semipermeable vesicle in a planar shear flow,
characterized by the elastic capillary number $Ca_\mathrm{E}=\chi\tau$,
where $\tau$ is the characteristic time scale in the
non-dimensionalization (see Supplementary Material) and $\chi$ is the
shear rate of the planar shear flow $\uu_{\infty}(\xx) = \chi (y,0)$.
When $\beta=0$ (impermeable case),  the vesicle reduced area and fluid
viscosity contrast determine whether a vesicle tank treads or
tumbles~\cite{fin-lam-sei-gom2008, kra-win-sei-lip1996}. Focusing on
semipermeable vesicles with no viscosity contrast, we find that a
semipermeable vesicle tilts to an inclination angle and undergoes
tank-treading dynamics, similar to the case of an impermeable vesicle.

In Figure~\ref{fig:shearComposite}(a), using four initial reduced areas
and two flow rates, we plot the reduced area of a semipermeable vesicle
with $\beta = 10^{-3}$. As observed in a quiescent flow, the final
vesicle shape is independent of the initial reduced area. However, it
does depend on the flow rate. We also observe that the amount of water inside the vesicle 
remains constant until $t\sim 10$ minutes, when 
%the vesicle starts to tank-tread and 
the normal component of the flow velocity 
becomes significantly smaller than the permeable velocity, leading to a
changing reduced area.

For a variety of dimensionless permeability coefficient $\beta>0$ and capillary number $Ca_\mathrm{E}$, a
semipermeable vesicle reaches an equilibrium shape as in
Figure~\ref{fig:shearComposite}(a). Since the equilibrium shape is selected by $Ca_\mathrm{E}$ and independent
of the initial reduced area, we initialize the vesicle with reduced area
$\alpha = 0.65$ for all simulations in
Figure~\ref{fig:shearComposite}(b), where we summarize the equilibrium
vesicle shape and its reduced area. We see that the capillary number
(flow rate) significantly affects the final vesicle shape, but the
permeability coefficient has a very minor effect. Instead, as we saw in the
quiescent analysis, $\beta$ sets a time scale since the time required
for a vesicle to reach its steady state configuration is proportional to
$\beta^{-1}$.

@@

At the beginning of the simulations a semipermeable vesicle is
placed above the center line $y=0$, and it migrates towards the
center line as in the impermeable case~\cite{dan-vla-mis2009}. 
%
%However,
%for an impermeable vesicle, depending on the initial reduced area and
%the maximum flow velocity $U$, the equilibrium vesicle shape is either
%an axisymmetric parachute or bullet, or an asymmetric tank-treading
%slipper~\cite{kao-bir-mis2009}. 
%
%To validate our methods,
%in Figure~\ref{fig:parabolicComposite}(a), we construct a phase diagram
%for the steady state shape of an impermeable vesicle
%vesicle~\cite{kao-bir-mis2009} (see Figure 2). While the vesicle shapes
%qualitatively agree, Kaoui et al.~report that the vesicles with
%$\alpha=0.7$ and flow rates $U = 1200 \mu m/s$ and $U = 1600 \mu m/s$ are
%axisymmetric. In our simulations these vesicles are slightly asymmetric,
%are undergoing a tank treading motion, and we therefore report them as
%asymmetric slippers.
%
Using several initial reduced areas and flow rates, we recreate the
impermeable vesicle phase diagram for vesicles with $\beta = 10^{-3}$.
The equilibrium vesicle shape for different $U$ and initial reduced area
is summarized in Figure~\ref{fig:parabolicComposite}(a), and the
vertical displacement of the vesicle's center of mass is in
Figure~\ref{fig:parabolicComposite}(b). The effect of semipermeability
of a vesicle with $\alpha(0) = 0.9$ and $U=800\; \mu$m/s on its reduced
area and migration pattern are shown in
Figures~\ref{fig:parabolicComposite}(c)
and~\ref{fig:parabolicComposite}(d), respectively.

As observed for the planar shear flow, the equilibrium vesicle reduced area is
independent of the initial reduced area and $\beta$, and  larger maximum flow velocity $U$ results in
smaller equilibrium reduced areas, as summarized in Figure~\ref{fig:parabolicComposite}(a). At
the smallest velocitty $U= 200\; \mu$m/s, the steady state reduced
area is large, and the equilibrium vesicle shape is an axisymmetric bullet. The equilibrium reduced area
decreases with increasing $U$, and the nonlinear flow profile gives rise to 
an asymmetric tank-treading slipper for $U\ge 400$ $\mu m/s$. We observe that the membrane
permeability to water drastically alters the equilibrium vesicle shape and position
relative to the Poiseuille flow: At $\beta=10^{-3}$ the equilibrium
vesicle depends only on the maximum velocity $U$ of the nonlinear shear
flow. The higher $U$ the smaller the equilibrium reduced area and the
further away vesicle is relative to the flow center at $y=0$.

@@

When cells go through small capillary vessels, they experience
large shape deformation and significant membrane stretching that leads to release of messengers 
%Cells experience stretching and release messengers when they pass
%through capillary vessels
~\cite{Wan2008_PNAS, ForsythWan2011_PNAS,Pak2015_PNAS}. 
Many cells exhibit extraordinary flexibility as they go
through narrow channels~\cite{AuStoreyMoore2016_PNAS}, and adjustment of
area to volume ratio may be a key factor for a successful passage.  Here
we quantify the effect of water permeability on a vesicle going through
a stenosis (a long microfluidic channel with a width smaller than the
cell radius).  Based on the findings in~\cite{abk-fai-sto2006}, we first
examine the effect of semipermeability on the pressure fluctuation
across a stenosed geometry. In this experiment, a pressure drop is used
to squeeze a red blood cell through a constriction with diameter
$5\;\mu$m and length $45\;\mu$m. We set up a similar computational
geometry with a Poiseuille flow imposed at the inlet and outlet
(Figure~\ref{fig:stenosisComposite}(a)). The flow velocity $U_{\max} =
1000 \;\mu$m/s at the inlet and outlet and is chosen so
that the residency time of the vesicle in the stenosis channel is $t_R
\approx 25$ ms which agrees with the experimental
results~\cite{abk-fai-sto2006}. When the red blood cell enters the
channel, an increase in the pressure drop is required to maintain a
constant flow rate, and the difference in the pressure drop is called
the excess pressure.
